---
title: "Mini Project #02"
author: "Aachal Ghimire"
date: "2024-10-13"
output: 
    html_document:
        code-fold: true
        code-summary: "show the code"
cache: true
---

Hi,

In this project, I am stepping into the role of a Hollywood development executive, where my job is to create new and exciting movie ideas. My goal is to develop fresh, data-driven concepts that bring something unique to the screen. To make that happen, I’ll be using IMDb data [the IMDb non- commercial release](https://datasets.imdbws.com/) to dig deep into Hollywood’s history, exploring patterns of success and failure in the industry.

With this data, we’ll analyze the track records of directors, actors, and genres to see what combinations have resonated with audiences. We’ll also look at metrics like box office earnings, critic and audience ratings, and even compare trends across decades to understand the factors that drive hits and avoid flops. By interpreting this information, I’ll gain valuable insights to guide the creation of movie concepts that have the potential to stand out in today’s market.

## **Libraries used**

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(readr)
```

## Data

```{r message=FALSE, warning=FALSE, cache=TRUE}

#IMDb_name_basics

get_imdb_file <- function(fname){
    BASE_URL <- "https://datasets.imdbws.com/"
    fname_ext <- paste0(fname, ".tsv.gz")
    if(!file.exists(fname_ext)){
        FILE_URL <- paste0(BASE_URL, fname_ext)
        download.file(FILE_URL, 
                      destfile = fname_ext)
    }
    as.data.frame(readr::read_tsv(fname_ext, lazy=FALSE))
}

NAME_BASICS <- read.csv("/Users/aachalghimire/Desktop/Fall/STA9750- matata668/MP#02.qmd/MP02/name_basics_small.csv", header = TRUE, sep = ",")


```

```{r  message=FALSE, warning=FALSE, cache=TRUE}
#IMDb_titile_basics

TITLE_BASICS     <- read.csv("/Users/aachalghimire/Desktop/Fall/STA9750- matata668/MP#02.qmd/MP02/title_basics_small.csv", header = TRUE, sep = ",")
```

```{r message=FALSE, warning=FALSE, cache=TRUE}
#IMDb_title_Episodes

TITLE_EPISODES   <- read.csv("/Users/aachalghimire/Desktop/Fall/STA9750- matata668/MP#02.qmd/MP02/title_episodes_small.csv", header = TRUE, sep = ",")
```

```{r message=FALSE, warning=FALSE, cache=TRUE}
#IMDb_title_ratings

TITLE_RATINGS    <- read.csv("/Users/aachalghimire/Desktop/Fall/STA9750- matata668/MP#02.qmd/MP02/title_ratings_small.csv", header = TRUE, sep = ",")
```

```{r message=FALSE, warning=FALSE, cache=TRUE}
#IMDb_title_crew

TITLE_CREW       <- read.csv("/Users/aachalghimire/Desktop/Fall/STA9750- matata668/MP#02.qmd/MP02/title_crew_small.csv", header = TRUE, sep = ",")
```

```{r message=FALSE, warning=FALSE, cache=TRUE}
#IMDb_title_principals

TITLE_PRINCIPALS <- read.csv("/Users/aachalghimire/Desktop/Fall/STA9750- matata668/MP#02.qmd/MP02/title_principals_small.csv", header = TRUE, sep = ",")

```

## Data Sub-Sampling

```{r}
library(dplyr)
library(stringr)
NAME_BASICS <- NAME_BASICS |>
    filter(str_count(knownForTitles, ",") > 1)
```

```{r}
TITLE_RATINGS |>
    ggplot(aes(x=numVotes)) + 
    geom_histogram(bins=30) +
    xlab("Number of IMDB Ratings") + 
    ylab("Number of Titles") + 
    ggtitle("Majority of IMDB Titles Have Less than 100 Ratings") + 
    theme_bw() + 
    scale_x_log10(label=scales::comma) + 
    scale_y_continuous(label=scales::comma)
```

```{r}
TITLE_RATINGS |>
    pull(numVotes) |>
    quantile()
```

```{r}
TITLE_RATINGS <- TITLE_RATINGS |>
    filter(numVotes >= 100)
```

## Semi_join

```{r}
TITLE_BASICS <- TITLE_BASICS |>
    semi_join(TITLE_RATINGS, 
              join_by(tconst == tconst))

TITLE_CREW <- TITLE_CREW |>
    semi_join(TITLE_RATINGS, 
              join_by(tconst == tconst))

TITLE_EPISODES_1 <- TITLE_EPISODES |>
    semi_join(TITLE_RATINGS, 
              join_by(tconst == tconst))
TITLE_EPISODES_2 <- TITLE_EPISODES |>
    semi_join(TITLE_RATINGS, 
              join_by(parentTconst == tconst))

TITLE_EPISODES <- bind_rows(TITLE_EPISODES_1,
                            TITLE_EPISODES_2) |>
    distinct()

TITLE_PRINCIPALS <- TITLE_PRINCIPALS |>
    semi_join(TITLE_RATINGS, join_by(tconst == tconst))


rm(TITLE_EPISODES_1)
rm(TITLE_EPISODES_2)
```

## Initial Exploration

```{r}
# Replace `\\N' with NA and then convert into numeric 
# glimpse() to view the characters 
NAME_BASICS|> glimpse()

NAME_BASICS_NA <- NAME_BASICS |>
    mutate(birthYear = as.numeric(replace(birthYear, birthYear == "\\N", NA)),
           deathYear = as.numeric(replace(deathYear, deathYear == "\\N", NA)))
```

## Initial Exploration

## Task 1
---

```{r warning=FALSE}
# replace "\\N" with NA 
# changed the charcters for the TITLE table using as.numeric and as.logical 
library(dplyr)


TITLE_BASICS_CCOL <- TITLE_BASICS|>
    mutate(
        isAdult = as.logical(isAdult), 
        startYear = as.numeric(replace(startYear, startYear == "\\N", NA)),
        endYear = as.numeric(replace(endYear, endYear == "\\N", NA)), 
        runtimeMinutes = as.numeric(runtimeMinutes))

TITLE_BASICS_CCOL |> glimpse()
print(TITLE_BASICS_CCOL)
View(TITLE_BASICS_CCOL)

#glimpse, offers a concise overview of a dataframe
# print (), outputs the contents of an object to the console
# view(),  opens a graphical spreadsheet
```

## Task 2
---

### 1.How many movies are in our data set? How many TV series? How many TV episodes?

```{r}
# filter
library(dplyr)
total_count <- TITLE_BASICS_CCOL |>
    filter(titleType %in% c("movie", "tvSeries", "tvEpisode")) |>
    group_by(titleType) |>
    summarize(count= n())

#Display the result

print(total_count)
```

### 2. Who is the oldest living person in our data set?

```{r}
#birthYear is not NA and deathYear is NA

oldest_living_person<- NAME_BASICS_NA |>
    filter(!is.na(birthYear) & is.na(deathYear)) |>
    slice_min(birthYear)

#birthYear>1900
oldest_living_person_1900 <- NAME_BASICS_NA |>
    filter(!is.na(birthYear) & is.na(deathYear), birthYear >=1900) |>
    slice_min(birthYear)

#age thresold is 110 years
#assign current year
#Filter and find the oldest person
current_year <- as.numeric(format(Sys.Date(), "%Y"))
year <- current_year - 110

oldest_living_person_110 <- NAME_BASICS_NA |>
    filter(!is.na(birthYear) & is.na(deathYear),birthYear >= year) |>
    arrange(birthYear) |>
    slice(1)


#display the result

print(oldest_living_person)
print(oldest_living_person_1900)
print(oldest_living_person_110)

```

We approached the question in three stages. First, analyzing the unfiltered data identified **Traudl Lessing (born 1625)** as the oldest living person, which is unlikely due to missing death years in the database. Next, we filtered for people born after 1900, which narrowed the list but still included entries with missing death years. Finally, we calculated a reasonable threshold by subtracting 110 years from the current year, identifying **Antonio Anelli (born 1914)** as the likely oldest living person, with a higher chance of being accurate.

### 3. There is one TV Episode in this data set with a perfect 10/10 rating and at least 200,000 IMDb ratings. What is it? What series does it belong to?

```{r}
#inner_join to join tables from TITLE_RATING, TITLE_EPISODES, TITLE_BASICS_CCOL
#select(-col) to drop unwanted columns 

perfect_rating <- TITLE_RATINGS |>
    filter(averageRating == 10, numVotes >= 200000) |>
    inner_join(TITLE_EPISODES, by = 'tconst') |>
    inner_join(TITLE_BASICS_CCOL, by = c("parentTconst" = 'tconst')) |>
    select(-isAdult, -startYear,-endYear, -runtimeMinutes, -genres )

#print result
print(perfect_rating)
```

**Breaking Bad season 5, episode 14** has the perfect rationg of 10/10 with **229313** rating

### 4. What four projects is the actor Mark Hamill most known for?

```{r}
#filter & identify the primary name = Mark Hamill
# filter and connect the data with tconst

NAME_BASICS_NA <- NAME_BASICS_NA|>
    mutate(primaryName = str_trim(primaryName))

Mark_Hamill_id <- NAME_BASICS_NA |>
    filter(primaryName == "Mark Hamill") |>
    select(nconst) |>
    pull()
   
#Filter his works in TITLE_PRINCIPAL

Mark_Hamill_works <- TITLE_PRINCIPALS |>
    filter(nconst == Mark_Hamill_id ) |>
    select(tconst)

# Select and join his top projects with ratings
 
Mark_Hamill_Projects <- Mark_Hamill_works |>
    inner_join(TITLE_BASICS_CCOL, by = "tconst") |>
    select(primaryTitle, tconst) |>
    inner_join(TITLE_RATINGS, by = "tconst") |>
    select(primaryTitle, averageRating, numVotes) |>
    arrange(desc(numVotes)) |>
    head(4)

print(Mark_Hamill_Projects)

```

Mark Hamill is widely known for his iconic role as Luke Skywalker in Star Wars. His top four projects, based on IMDb’s highest number of user votes, are Star Wars: Episode IV - A New Hope, Episode V - The Empire Strikes Back, Episode VI - Return of the Jedi, and Episode VII - The Force Awakens. These movies highlight his most celebrated contributions to the series and cinema.

### 5. What TV series, with more than 12 episodes, has the highest average rating?

```{r}

#calculating episodes for each series and filter >12

total_episodes <- TITLE_EPISODES |>
    group_by(parentTconst) |>
    summarise(total_episodes = n()) |>
    filter(total_episodes > 12)

#making sure there are series with more than 12 eps
print(total_episodes)

# Ensuring column types match and remove missing values if necessary
TITLE_BASICS_CCOL <- TITLE_BASICS_CCOL |>
    mutate(tconst = as.character(tconst))

total_episodes <- total_episodes |>
    mutate(parentTconst = as.character(parentTconst))

# filter tvSeries and join with total episodes

series_episodes <- TITLE_BASICS_CCOL |>
    filter(titleType == "tvSeries") |>
    inner_join(total_episodes, by = c("tconst" = "parentTconst"))

print(series_episodes)

#Joining with rating and finding the highest rated series

TVSeries_12eps_rating <- series_episodes |>
    inner_join(TITLE_RATINGS, by = "tconst") |>
    arrange(desc(averageRating)) |>
    select(primaryTitle, total_episodes, averageRating) |>
    slice(1)

print(TVSeries_12eps_rating)

```

**Craft Games** among tvShows with total of **318 episodes** has the highest rating of **9.7**

### 6. The TV series Happy Days (1974-1984) gives us the common idiom “jump the shark”. The phrase comes from a controversial fifth season episode (aired in 1977) in which a lead character literally jumped over a shark on water skis. Idiomatically, it is used to refer to the moment when a once-great show becomes ridiculous and rapidly looses quality.

Is it true that episodes from later seasons of Happy Days have lower average ratings than the early seasons?

```{r}

#calculating all the episodes of Happy Days

Happy_days_season <- TITLE_BASICS_CCOL |>
    filter(primaryTitle == "Happy Days", titleType == "tvSeries") |>
    select(tconst) |>
    pull()

#episodes by season number

HappyDays_episodes <- TITLE_EPISODES |>
    filter(parentTconst %in% Happy_days_season) |>
    mutate(seasonCategory = ifelse(seasonNumber <= 4, "early", "later")) 

HappyDays_rating <- HappyDays_episodes |>
    inner_join(TITLE_RATINGS, by= 'tconst') |>
    group_by(seasonCategory) |>
    summarize(averageRating = mean(averageRating, na.rm = TRUE))

print(HappyDays_rating)

```

Based on the analysis, we observe a decline in average ratings from early to later seasons. The early seasons had an average rating of 7.58, whereas the later seasons dropped to 6.91. This indicates a noticeable decrease in viewer ratings as the series progressed.


## Quantifying Success

## Task 3: custom success matrix
---

```{r}
#adding success rating to TITLE_RATING column

TITLE_RATING_SUCCESS <- TITLE_RATINGS |>
    mutate(success = averageRating * log(numVotes)) 

print(TITLE_RATING_SUCCESS)

```

### 1. Choose the top 5-10 movies on your metric and confirm that they were indeed box office successes.

```{r}
top_sucessful_movies <- TITLE_RATING_SUCCESS |>
    arrange(desc(success)) |>
    slice(1:10)

print(top_sucessful_movies)
```

top 10 successful movies using our success matrix.

### 2. Choose 3-5 movies with large numbers of IMDb votes that score poorly on your success metric and confirm that they are indeed of low quality.

```{r}
low_successful_movies <- TITLE_RATING_SUCCESS |>
    filter(numVotes >= 50000) |>
    arrange(success) |>
    slice(1:5)
print(low_successful_movies)

```

Our analysis reveals a list of movies that, despite being popular, have low ratings. Cross-referencing reviews on platforms like Google confirms that these movies are generally considered low quality.

### 3. Choose a prestige actor or director and confirm that they have many projects with high scores on your success metric.

```{r}
# we have chosen Zendaya
actor_id <- NAME_BASICS_NA |>
    filter(primaryName == "Zendaya") |>
    select(nconst) |>
    pull()

#filter for Zendaya's projects 

actor_projects <- TITLE_PRINCIPALS |>
    filter(nconst == actor_id) |>
    select(tconst)

# join the project table with rating to measure the success

actor_success <- actor_projects |>
    inner_join(TITLE_BASICS_CCOL, by = 'tconst') |>
    inner_join(TITLE_RATING_SUCCESS, by = 'tconst') |>
    arrange(desc(success)) |>
    select(primaryTitle, averageRating, numVotes, success)

#print the result
print(actor_success)

```

Our analysis confirms that Zendaya has a successful portfolio with multiple highly rated projects that perform strongly on our success metric, reflecting both high quality and broad popularity. 
    
    
### 4. Perform at least one other form of ‘spot check’ validation.

```{r}
#top 3 movie bassed on success matrix
#bottom 3 movies based on success matrix

top_3movies <- TITLE_RATING_SUCCESS |>
    arrange(desc(success)) |>
    slice_head( n = 3)

bottom_3movies <- TITLE_RATING_SUCCESS |>
    arrange(success) |>
    slice_head(n = 3)

#combinig two tabels for a unified check

spot_check_movies <- bind_rows(top_3movies, bottom_3movies)
print(spot_check_movies)
```

Our spot check confirmed that the success metric aligns well with IMDb ratings, accurately identifying both high- and low-rated movies, which shows that our analysis is reliable.


### 5. Come up with a numerical threshold for a project to be a ‘success’; that is, determine a value v such that movies above v are all “solid” or better. 

```{r}
# Load necessary libraries
library(dplyr)

# calculate based on averageRating and numVotes
TITLE_RATING_SUCCESS2 <- TITLE_RATING_SUCCESS |>
    mutate(success = averageRating * log(numVotes + 1))

# Calculate the threshold (75th percentile) for success, fixing the typo to 'threshold'
threshold <- quantile(TITLE_RATING_SUCCESS2$success, 0.75, na.rm = TRUE)

# Filter for successful movies with success >= threshold and averageRating >= 7
successful_movies <- TITLE_RATING_SUCCESS2 |>
    filter(success >= threshold, averageRating >= 7)

# Display the threshold and successful movies
print(paste("Success threshold (75th percentile):", threshold))
print(successful_movies)

```
Based on our analysis and the calculated threshold, the 75th percentile for the success metric is **49.0371**.

## Examining Success. by Genre and Decade
## Task 4
---

### 1.  What was the genre with the most “successes” in each decade?

```{r}
# Filter for movies in title_basics and title_rating

movie_ratings <- TITLE_BASICS_CCOL|>
    inner_join(TITLE_RATING_SUCCESS, by= 'tconst')

#new colum for decade

decade_movie_rating <- movie_ratings |>
    mutate(decade = floor(startYear/10)*10) |>
    filter(decade >= 1980)

# group by decade and genre

decade_genre <- decade_movie_rating |>
    group_by(decade, genres) |>
    summarize(total_success = sum(success, na.rm = TRUE), .groups = 'drop')
    
most_successful_by_decade_genre <- decade_genre |>
  group_by(decade) |>
  slice_max(order_by = total_success, n = 1)


#Display the result

print(most_successful_by_decade_genre)
```
From our analysis, we can conclude that **Comedy** was the most popular genre from the 1980s onwards. However, since 2020, Drama has emerged as the leading genre. It will be interesting to see which genre dominates in the next decade.

## 2. What genre consistently has the most “successes”? What genre used to reliably produced “successes” and has fallen out of favor?

```{r}
# number of times each genre was the most successful across all decades
constant_genre_success <- most_successful_by_decade_genre |>
  group_by(genres) |>
  summarize(count = n()) |>
  arrange(desc(count))

# Display the genre that has the most consistent successes
print(constant_genre_success)

# Filter for genres that were successful in early decades but not in the latest ones
fallen_genres <- most_successful_by_decade_genre |>
  filter(decade < 2000) |>
  anti_join(most_successful_by_decade_genre %>% filter(decade >= 2000), by = "genres")

# Display the genres that have fallen out of favor
print(fallen_genres)


```



## Successful Personnel in the Genre

## Task 5
---

Identify (at least) two actors and one director who you will target as the key talent for your movie. Write a short “pitch” as to why they are likely to be successful. You should support your pitch with at least one graphic and one table.

```{r}








